Staged Abstract Interpreters

### 2019-04-05 paper/10_intro.tex
paper

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

\begin{itemize}
\item Monadic abstractions, which encapsulate the following two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\end{itemize}

We show that these well-known abstractions can be combined smoothly, and more
importantly the price is almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. \todo{put some numbers} However, we do not lose any advantages
of this high-level programming style (e.g., modularity and extensibility); the
meaning of the program also remains the same, particularly, the abstract
semantics that does the analysis when instantiated as an abstract interpreter.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not a new, despite the fact that most of the previous work
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation from. In this paper, we advocate a systematic
approach to meta-programming that achieves the performance goal with less
intrusive changes; in fact, a case study (Section~\ref{cs_ac}) comparing our
approach with abstract compilation shows that, by utilizing type-based stage
annotations, the analyzer program does not need any changes.

@@

\paragraph{Contributions} The contribution of the paper is twofold:
1) Intellectually, we present a unifying construction that naturally extends the first 
   Futamura projection to abstract interpreters.
2) Pragmatically, we show that staging abstract interpreters is an approach to
  improve performance while untouch the soundness by case studies and emprical
  evaluation. The paper is organized as follows:

\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{generic_if}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with least engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    \note{garbage collection}, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive, \note{context-sensitive, and store-widening} analysis.
\end{itemize}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) 
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

\begin{itemize}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\item Monadic abstractions, which encapsulate the above two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\end{itemize}

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. \todo{put some numbers} However, we do not lose any advantages
of this high-level programming style (e.g., modularity and extensibility); the
meaning of the program also remains the same, particularly, the soundness of 
static analysis after staging.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous work
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach to meta-programming that achieves the performance goal with less
intrusive changes; in fact, a case study (Section~\ref{cs_ac}) comparing our
approach with abstract compilation shows that, by utilizing type-based stage
annotations, the analyzer program does not need any changes.

@@

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and emprical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouch 
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    \note{garbage collection}, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive, \note{context-sensitive, and store-widening} analysis.
\end{itemize}

### 2019-04-05 paper/20_prelim.tex
paper

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monad to demonstrate the
idea, but the approach is not restricted to our choice. One can use
imperative/direct style and other MSP languages (e.g., MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} and Template
Haskell \cite{Sheard:2002:TMH:636517.636528} to construct such staged abstract
interpreters.

@@

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions and conditionals. Other effectful features
such as assignments can also be supported readily. In Section~\ref{cases_imp}, we
will add more imperative features to the language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguised any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

@@

The abstract syntax we present in fact can be seen as a deep embedding of the
language -- we use data-types to represent programs. This design choice makes us
easily add different interpretations over the AST; with the inheritance and
overriding mechanism in Scala, we may also add new language construct and reuse
the existing interpretation.

@@

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
takes a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one. Constructing
extensible interpreters using monad transformers was first proposed by
\citet{DBLP:conf/popl/LiangHJ95}, and later has been applied to abstract
interpreters \cite{Sergey:2013:MAI:2491956.2491979,
DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}. In section,
with the multi-stage programming and monad transformers in mind while leave them
as abstract type members, we present the generic interface of a big-step
definitional interpreter.

@@

We also defines several methods to obtain and modify the environment and store.
These methods either returns monadic values with environment or store inside, or
simply a @Unit@ value for effects. For example, @local_env@ takes a value @ans@ of
type @Ans@ and a new environment which will be installed when evaluating @ans@.

@@

Here we sketch the basic idea of unstaged @ReaderT@ and @StateT@. Readers may
refer to \todo{cite, or appendix?} for more detail and implementation of monad
transformers.
A @ReaderT@ monad transformer encapsulates the computation @R => M[A]@, where
@R@ is the reader type, @M[_]@ is a monad type. The @ReaderT@ monad returns the
transformed value of type @M[A]@. Similarly, a @StateT@ monad encapsulates the
computation @S => M[(A, S)]@, where @S@ is the state type, @M[_]@ is a monad
type. The @StateT@ monad returns the transformed value of type @M[(A, S)]@,
where the new state is accompanied.
Note that the binding-time type @R@ is an identity type, thus the monads are
operated on unstaged data. We can also see this from the signature of @flatMap@:
the function @f@ takes an unstaged value of type @A@ and produces a monadic
value.

>>>>

In this section, we first describe the abstract syntax of the language for our interpreter, 
then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monad to demonstrate the
idea, but the approach is not restricted to our choice. One can use
imperative/direct style and other MSP languages (e.g., MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} and Template
Haskell \cite{Sheard:2002:TMH:636517.636528} to construct such staged abstract
interpreters.

@@

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions and conditionals. Other effectful features
such as assignments can also be supported readily. In Section~\ref{cases_imp}, we
will add more imperative features to the language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguise any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

@@

The abstract syntax we present in fact can be seen as a deep embedding of the
language -- we use data-types to represent programs. This design choice allows us to
easily use different interpretations over the AST; with the inheritance and
overriding mechanism in Scala, we may also add new language construct and reuse
the existing interpretation.

@@

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
take a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one. Constructing
extensible interpreters using monad transformers was first proposed by
\citet{DBLP:conf/popl/LiangHJ95}, and later has been applied to abstract
interpreters \cite{Sergey:2013:MAI:2491956.2491979,
DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}. In this section,
with the multi-stage programming and monad transformers in mind while leaving them
as abstract type members, we present the generic interface of a big-step
definitional interpreter.

@@

We also defines several methods to obtain and modify the environment and store.
These methods return monadic values of type @AnsM[_]@, which encapsulate the 
environment or store, or simply a @Unit@ value for effects.
For example, @local_env@ takes a value @ans@ of type @Ans@ and an environment 
which will be installed when evaluating @ans@.

@@

Here we sketch the basic idea of unstaged @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail 
and implementation of monad transformers.
A @ReaderT@ monad transformer encapsulates the computation @R => M[A]@, where
@R@ is the reader type, @M[_]@ is a monad type. The @ReaderT@ monad returns the
transformed value of type @M[A]@. Similarly, a @StateT@ monad encapsulates the
computation @S => M[(A, S)]@, where @S@ is the state type, @M[_]@ is a monad
type. The @StateT@ monad returns the transformed value of type @M[(A, S)]@,
where the new state is accompanied.
Note that the binding-time type @R@ is an identity type, thus the monads are
operated on unstaged data. We can also see this from the signature of @flatMap@:
the function @f@ takes an unstaged value of type @A@ and produces a monadic
value.

### 2019-04-05 paper/70_eval.tex
misc
paper

We use Scala 2.12.4 and Oracle JVM 1.8.0-191, running on an Ubuntu 16.04 LTS
(kernel 4.4.0) machine. All of our evaluation were performed on a machine with 4 Intel
Xeon Platinum 8168 CPU at 2.7GHz and 3 TiB of RAM. Although the machine has 96
cores (192 threads) in total, the abstract interpreters are single-threaded.
To minimize the effect of warming-up from the JVM JIT compiler, all the
experiments are executed for 20 times and we report the statistical median value
of the running times (in seconds).

@@

\begin{figure}[h]
\tiny
\begin{tabular}{@{}lllll@{}}
\toprule
    Program          &\#AST & unstaged   & staged     & $\Delta$ \\ \midrule
    @rsa@            &      &            &            & \%          \\
    @kcfa3@          &      &            &            & \%          \\
    @church@         &      &            &            & \%          \\
    @fermat@         &      &            &            & \%          \\
    @kcfa-worst-16@  &  198 & 0.352      & 0.028      & +1157.14\%        \\
    @kcfa-worst-32@  &  390 & 4.234      & 0.289      & +1365.05\%          \\
    @kcfa-worst-64@  &  774 & 61.475     & 4.425      & +1289.27\%         \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for 0CFA without store-widening.} \label{evaluation_result}
\end{figure}

>>>>

We use Scala 2.12.7 and Oracle JVM 1.8.0-191, running on an Ubuntu 16.04 LTS
(kernel 4.4.0) machine. All of our evaluation were performed on a machine with 4 Intel
Xeon Platinum 8168 CPU at 2.7GHz and 3 TiB of RAM. Although the machine has 96
cores (192 threads) in total, the abstract interpreters are single-threaded.
To minimize the effect of warming-up from the JVM JIT compiler, all the
experiments are executed for 20 times and we report the statistical median value
of the running times (in seconds).

@@

\begin{figure}[h]
\tiny
\begin{tabular}{@{}lllll@{}}
\toprule
    program          &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ \\ \midrule
    @church@         & 120  & 71.811     & 23.990     & 299.34\%          \\
    @rsa@            & 451  & 0.308      & 0.064      & 459.15\%          \\
    @latice@         & 609  & 31.874     & 8.816      & 358.69\%          \\
    @fermat@         & 310  & 0.043      & 0.006      & 716.67\%          \\
    @mbrotZ@         & 331  & 1.112      & 0.214      & 519.63\%          \\
    @kcfa-worst-16@  & 182  & 0.083      & 0.013      & 638.46\%          \\
    @kcfa-worst-32@  & 358  & 0.452      & 0.044      & 1027.27\%         \\
    @kcfa-worst-64@  & 774  & 3.329      & 0.236      & 1410.59\%         \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for 0CFA without store-widening.} \label{evaluation_result}
\end{figure}

### 2019-04-05 paper/abstract.tex
paper

It is well known that a staged interpreter is a compiler: specializing the
interpreter to a given program produces an equivalent program that runs faster,
which known as the first Futamura projection. It is even more widely known that an
abstract interpreter is a program analyzer: tweaking the interpreter to run on a
domain of abstract values produces a sound static analysis. What happens when we
combine these two ideas, and apply staging to an \emph{abstract} interpreter?

>>>>

It is well known that a staged interpreter is a compiler: specializing the
interpreter to a given program produces an equivalent program that runs faster,
which is known as the first Futamura projection. It is even more widely known that an
abstract interpreter is a program analyzer: tweaking the interpreter to run on a
domain of abstract values produces a sound static analysis. What happens when we
combine these two ideas, and apply staging to an \emph{abstract} interpreter?
